Sparse Representation Approaches for First Stage Retrieval: A Review

### 2021-12-19 src/article.tex
Update SPLADE loss function.

    \paragraph{Training the importance predictor.}
    There is a set of N training instances $R=\{(q_1, d_{1+}, d_{1-}), ..., (q_N, d_N+, d_N-)\}$,
    each of which contains a query $q_i$, a positive document (passage) $d_{i+}$ and a negative
    one $d_i-$, such that the first document is more relevant the query than the second one.
    As a result, the loss function has the following form:
    \begin{equation}
        L_{rank} = -log \frac
        {e^{s(q_i,d_{i+})}}
        {e^{s(q_i,d_{i+})} + e^{s(q_i,d_{i-})}}
    \end{equation}
    where $s(q,d)$ is denote the ranking score obtained via dot product between $d$ and $q$
    representations from Equation (\ref{eqn:sparterm:mapping-process}).

@@

    \paragraph{Architecture}
    \begin{figure}[h]
        \centering
        \def\svgwidth{\columnwidth}
        \input{SPLADE.pdf_tex}
        \caption{Final representation calculation with log saturation}
        \label{fig:SPLADE}
    \end{figure}
    The importance estimation formula Equation (\ref{eqn:sparterm:importance-predictor})
    has been modified by adding a log-saturation effect (Figure \ref{fig:SPLADE}), in order to
    prevent the dominance of some terms to dominate and naturally ensures sparsity in
    representations:
    \begin{equation}
        w_j=\sum_{i \in t}log(1+ReLU(w_{ij}))
    \end{equation}

    \paragraph{Loss function}
    The loss function has the following form:
    \begin{equation}
        L_{rank-IBN}=-log
        \frac{
            e^{s(q_i,d_i^+)}
        }{
            e^{s(q_i,d_i^+)} +
            e^{s(q_i,d_i^-)} +
            \sum_j e^{s(q_i,d_{i,j}^-)}
        }
    \end{equation}
    where IBN is in-batch negatives sampling strategy, which is effective for learning first-stage
    rankers \cite{IBN1,IBN2,IBN3}, $q_i$ is query, $d_i^+$ and $d_i^-$ are positive and negative
    documents respectively, ${d_{i,j}^-}$ is set of positive documents from other queries,
    but negative for the current one \cite{SPLADE}.\\\par
    As a result, SPLADE is simple and effective model in information retrieval tasks.

>>>>

    \paragraph{Training the importance predictor.}
    There is a set of N training instances $R=\{(q_1, d_{1+}, d_{1-}), ..., (q_N, d_N+, d_N-)\}$,
    each of which contains a query $q_i$, a positive document (passage) $d_{i+}$ and a negative
    one $d_i-$, such that the first document is more relevant the query than the second one.
    As a result, the loss function has the following form:
    \begin{equation}
        \label{eqn:sparterm:loss}
        L_{rank} = -log \frac
        {e^{s(q_i,d_{i+})}}
        {e^{s(q_i,d_{i+})} + e^{s(q_i,d_{i-})}}
    \end{equation}
    where $s(q,d)$ is denote the ranking score obtained via dot product between $d$ and $q$
    representations from Equation (\ref{eqn:sparterm:mapping-process}).

@@

    \paragraph{Architecture}
    \begin{figure}[h]
        \centering
        \def\svgwidth{\columnwidth}
        \input{SPLADE.pdf_tex}
        \caption{Final representation calculation with log saturation}
        \label{fig:SPLADE}
    \end{figure}
    The importance estimation formula Equation (\ref{eqn:sparterm:importance-predictor})
    has been modified by adding a log-saturation effect (Figure \ref{fig:SPLADE}), in order to
    prevent the dominance of some terms to dominate and naturally ensures sparsity in
    representations:
    \begin{equation}
        \label{eqn:splade:importance}
        w_j=\sum_{i \in t}log(1+ReLU(w_{ij}))
    \end{equation}

    \paragraph{Loss function}
    The function presented in the Equation (\ref{eqn:sparterm:loss}) is supplemented with negative
    sampling as shown below:
    \begin{equation}
        L_{rank-IBN}=-log
        \frac{
            e^{s(q_i,d_i^+)}
        }{
            e^{s(q_i,d_i^+)} +
            e^{s(q_i,d_i^-)} +
            \sum_j e^{s(q_i,d_{i,j}^-)}
        }
    \end{equation}
    where IBN is in-batch negatives sampling strategy, which is effective for learning first-stage
    rankers \cite{IBN1,IBN2,IBN3}, $q_i$ is query, $d_i^+$ and $d_i^-$ are positive and negative
    documents respectively, ${d_{i,j}^-}$ is set of positive documents from other queries,
    but negative for the current one \cite{SPLADE}.

    This approach does not use binary gating as in SparTerm, and instead learn model (Equation
    \ref{eqn:splade:importance}) jointly optimizing ranking and regularization losses:
    \begin{equation}
        L = L_{rank-IBN} + \lambda_q L^q_{reg} + \lambda_d L^d_{reg}
    \end{equation}
    where $L_{reg}$ is a sparse regularisation ($L_1$ or $L_{FLOPS}$), $\lambda_q$ and
    $\lambda_d$ are tunable hyper-parameters to control the sparsity of the query and the
    document. \newline

    As a result, SPLADE is simple and effective model in information retrieval tasks.
